---
title: "Seasonality Detection Methods: A Comparative Study"
author: "fdars Package"
date: "2025-12-29"
format:
  pdf:
    pdf-engine: pdflatex
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    geometry:
      - margin=0.75in
    fig-width: 7
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

This report describes and compares five methods for detecting seasonality in functional time series data. We evaluate each method's performance across different scenarios including varying seasonal strengths, non-linear trends, and different trend types.

The goal is to answer: **Given a time series, how can we reliably determine if it contains a seasonal pattern?**

# Detection Methods

## AIC Comparison (Fourier vs P-spline)

**Concept**: If data is seasonal, a Fourier basis should fit better than P-splines because Fourier bases naturally capture periodic patterns.

**Mathematical formulation**:

For a curve $y(t)$, we fit two models:

1. **Fourier basis**: $\hat{y}(t) = \sum_{k=0}^{K} a_k \cos(2\pi k t) + b_k \sin(2\pi k t)$

2. **P-spline**: $\hat{y}(t) = \sum_{j=1}^{J} c_j B_j(t)$ with penalty $\lambda \int [\hat{y}''(t)]^2 dt$

We compute AIC for each:
$$\text{AIC} = n \log(\text{RSS}/n) + 2 \cdot \text{edf}$$

where RSS is the residual sum of squares and edf is the effective degrees of freedom.

**Detection rule**: Seasonality detected if $\text{AIC}_{\text{P-spline}} - \text{AIC}_{\text{Fourier}} > 0$

**Interpretation**: When Fourier has lower AIC, the periodic structure is significant enough to justify the global periodic assumption over the local flexibility of splines.

## FFT Confidence

**Concept**: Use Fast Fourier Transform to detect dominant frequencies. Strong peaks in the periodogram indicate periodic components.

**Mathematical formulation**:

Given a time series $y_1, y_2, \ldots, y_n$, compute the discrete Fourier transform:
$$Y_k = \sum_{j=1}^{n} y_j e^{-2\pi i (j-1)(k-1)/n}$$

The periodogram (power spectrum) is:
$$P_k = |Y_k|^2$$

**Detection score**:
$$\text{Confidence} = \frac{\max_k P_k}{\text{mean}(P_k)}$$

**Detection rule**: Seasonality detected if Confidence $> 6.0$

**Interpretation**: A high ratio indicates one frequency dominates, suggesting periodicity rather than random noise.

## ACF Confidence

**Concept**: Autocorrelation at the seasonal lag should be high for seasonal data.

**Mathematical formulation**:

The autocorrelation function at lag $h$ is:
$$\rho_h = \frac{\sum_{t=1}^{n-h}(y_t - \bar{y})(y_{t+h} - \bar{y})}{\sum_{t=1}^{n}(y_t - \bar{y})^2}$$

For seasonal data with period $p$, we expect $\rho_p$ to be significantly positive.

**Detection rule**: Seasonality detected if ACF confidence $> 0.25$

**Interpretation**: High autocorrelation at the seasonal lag indicates the pattern repeats.

## Variance Strength

**Concept**: Decompose variance into seasonal and residual components. High seasonal variance ratio indicates seasonality.

**Mathematical formulation**:

Decompose the series: $y_t = T_t + S_t + R_t$ (trend + seasonal + residual)

The seasonal strength is:
$$\text{SS}_{\text{var}} = 1 - \frac{\text{Var}(R_t)}{\text{Var}(y_t - T_t)}$$

**Detection rule**: Seasonality detected if $\text{SS}_{\text{var}} > 0.2$

**Interpretation**: Values close to 1 mean the seasonal component dominates; values close to 0 mean residual noise dominates.

**Important**: The `period` parameter must be in the same units as `argvals`. For data normalized to [0,1] with 5 annual cycles, use `period = 0.2`.

## Spectral Strength

**Concept**: Measure the proportion of spectral power at the seasonal frequency.

**Mathematical formulation**:

Using the periodogram $P_k$, identify the seasonal frequency $f_s = 1/\text{period}$.

$$\text{SS}_{\text{spectral}} = \frac{\sum_{k \in \mathcal{S}} P_k}{\sum_{k} P_k}$$

where $\mathcal{S}$ includes the seasonal frequency and its harmonics.

**Detection rule**: Seasonality detected if $\text{SS}_{\text{spectral}} > 0.3$

**Interpretation**: High values indicate spectral energy is concentrated at seasonal frequencies.

# Simulation Studies

## Simulation 1: Varying Seasonal Strength {#sec-sim1}

### Setup

This simulation tests how well each method detects seasonality at different signal strengths.

**Parameters**:

- 11 seasonal strength levels: 0.0, 0.1, ..., 1.0
- 50 curves per strength level
- 5 years of monthly data (60 observations)
- Noise standard deviation: 0.3

**Signal model**:
$$y(t) = s \cdot [\sin(2\pi \cdot 5t) + 0.3\cos(4\pi \cdot 5t)] + \epsilon, \quad \epsilon \sim N(0, 0.3^2)$$

where $s$ is the seasonal strength (0 = no seasonality, 1 = full seasonality).

**Ground truth**: A curve is classified as "truly seasonal" if $s \geq 0.2$.

### Code

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show simulation code"

library(fdars)
library(ggplot2)
library(tidyr)
library(dplyr)

set.seed(42)

# Configuration
n_strengths <- 11
n_curves_per_strength <- 50
n_years <- 5
n_months <- n_years * 12
noise_sd <- 0.3

# Detection thresholds (calibrated to ~5% FPR on pure noise)
detection_thresholds <- list(
  aic_comparison = 0,
  fft_confidence = 6.0,
  acf_confidence = 0.25,
  strength_variance = 0.2,
  strength_spectral = 0.3
)

seasonal_strengths <- seq(0, 1, length.out = n_strengths)
t <- seq(0, 1, length.out = n_months)

# Generate seasonal curve
generate_seasonal_curve <- function(t, strength, noise_sd = 0.3) {
  n_cycles <- length(t) / 12
  seasonal <- strength * sin(2 * pi * n_cycles * t)
  seasonal <- seasonal + strength * 0.3 * cos(4 * pi * n_cycles * t)
  noise <- rnorm(length(t), sd = noise_sd)
  return(seasonal + noise)
}
```

### Results

![Detection rates by seasonal strength](plots/seasonality_detection_comparison.pdf){#fig-detection-rates width="7in"}

\newpage

**How to interpret**:

- The x-axis shows the true seasonal strength (0 = pure noise, 1 = strong seasonality)
- The y-axis shows what percentage of curves each method classified as "seasonal"
- The vertical dashed line at 0.2 marks the ground truth threshold
- **Ideal behavior**: 0% detection below the threshold, 100% above

### Classification Performance

| Method | F1 Score | Precision | Recall | FPR | Specificity |
|--------|----------|-----------|--------|-----|-------------|
| **Variance Strength** | **97.3%** | 98.2% | 96.4% | 2.0% | 92.0% |
| Spectral Strength | 95.3% | 97.4% | 93.3% | 10.0% | 89.0% |
| FFT Confidence | 94.8% | 99.3% | 90.7% | 4.0% | 97.0% |
| AIC Comparison | 91.5% | 94.3% | 88.9% | 18.0% | 76.0% |
| ACF Confidence | 85.4% | 98.3% | 75.6% | 10.0% | 94.0% |

**How to interpret**:

- **F1 Score**: Harmonic mean of precision and recall (higher is better)
- **Precision**: Of curves detected as seasonal, what % are truly seasonal?
- **Recall**: Of truly seasonal curves, what % did we detect?
- **FPR**: False Positive Rate - what % of non-seasonal curves were incorrectly flagged?

### Precision-Recall Analysis

![Precision-Recall curves](plots/seasonality_detection_details.pdf){#fig-pr-curves width="7in"}

\newpage

**How to interpret**:

- Curves closer to the top-right corner are better
- The diamond markers show the operating point at the default threshold
- AUC-PR (Area Under the PR Curve) summarizes overall performance

## Simulation 2: Non-linear Trend {#sec-sim2}

### Setup

This simulation tests robustness when non-linear trends are added to the seasonal signal.

**Parameters**:

- 6 seasonal strength levels × 6 trend strength levels
- 30 curves per combination
- Non-linear trend: quadratic + cubic + sigmoid components

**Signal model**:
$$y(t) = \text{Trend}(t, \tau) + \text{Seasonal}(t, s) + \epsilon$$

where $\tau$ is the trend strength and $s$ is the seasonal strength.

**Trend function**:
$$\text{Trend}(t, \tau) = \tau \cdot [2(t-0.5)^2 + 0.5(t-0.3)^3 + 0.3 \cdot \sigma(10(t-0.6)) - 0.5]$$

### Code

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show trend simulation code"

# Non-linear trend function
generate_nonlinear_trend <- function(t, trend_strength) {
  quadratic <- 2 * (t - 0.5)^2
  cubic <- 0.5 * (t - 0.3)^3
  sigmoid <- 1 / (1 + exp(-10 * (t - 0.6)))
  trend <- trend_strength * (quadratic + cubic + 0.3 * sigmoid - 0.5)
  return(trend)
}

# Generate curve with trend + seasonal + noise
generate_curve <- function(t, seasonal_strength, trend_strength, noise_sd = 0.3) {
  trend <- generate_nonlinear_trend(t, trend_strength)
  n_cycles <- length(t) / 12
  seasonal <- seasonal_strength * sin(2 * pi * n_cycles * t)
  seasonal <- seasonal + seasonal_strength * 0.3 * cos(4 * pi * n_cycles * t)
  noise <- rnorm(length(t), sd = noise_sd)
  return(trend + seasonal + noise)
}
```

### Results

![Detection rates heatmap by seasonal and trend strength](plots/seasonality_detection_trend_heatmaps.pdf){#fig-trend-heatmap width="7in"}

\newpage

**How to interpret**:

- Each cell shows the detection rate for a combination of seasonal strength (x) and trend strength (y)
- Blue = low detection rate, Red = high detection rate
- The dashed line separates non-seasonal (left) from seasonal (right) ground truth

### F1 Score vs Trend Strength

| Method | No Trend | Max Trend | F1 Drop |
|--------|----------|-----------|---------|
| Spectral | 96.3% | 92.5% | 3.9% |
| FFT | 93.7% | 91.8% | 2.0% |
| AIC | 92.2% | 87.0% | 5.7% |
| ACF | 87.4% | 83.5% | 4.5% |

**How to interpret**:

- **F1 Drop**: How much performance degrades when strong trends are present
- Lower drop = more robust to trends

### False Positive Rate by Trend Strength

![FPR when no seasonality is present, across trend strengths](plots/seasonality_detection_trend_pr.pdf){#fig-trend-fpr width="7in"}

**Key finding**: FFT's FPR increases dramatically with trend strength because non-linear trends can create spurious peaks in the periodogram.

## Simulation 3: Multiple Trend Types {#sec-sim3}

### Setup

This simulation tests which types of trends cause the most problems for each detection method.

**Trend types tested**:

1. **None**: Flat baseline
2. **Linear**: $f(t) = t - 0.5$
3. **Quadratic**: $f(t) = (t-0.5)^2 - 0.25$
4. **Cubic**: $f(t) = 2(t-0.5)^3$
5. **Exponential**: $f(t) = e^{2t}/e^2 - 0.5$
6. **Logarithmic**: $f(t) = \log(t+0.1)$ (normalized)
7. **Sigmoid**: $f(t) = 1/(1+e^{-10(t-0.5)}) - 0.5$
8. **Slow sine**: $f(t) = \sin(2\pi t)$ — one cycle over the entire series

### Code

```{r}
#| eval: false
#| code-fold: true
#| code-summary: "Show trend types code"

trend_functions <- list(
  none = function(t, strength) rep(0, length(t)),
  linear = function(t, strength) strength * (t - 0.5),
  quadratic = function(t, strength) strength * ((t - 0.5)^2 - 0.25),
  cubic = function(t, strength) strength * 2 * (t - 0.5)^3,
  exponential = function(t, strength) strength * (exp(2 * t) / exp(2) - 0.5),
  logarithmic = function(t, strength) {
    strength * (log(t + 0.1) - log(0.1)) / (log(1.1) - log(0.1)) - 0.5 * strength
  },
  sigmoid = function(t, strength) strength * (1 / (1 + exp(-10 * (t - 0.5))) - 0.5),
  slow_sine = function(t, strength) strength * sin(2 * pi * t)
)
```

### Results

![F1 scores by trend type](plots/seasonality_detection_trend_types_f1.pdf){#fig-trend-types-f1 width="7in"}

\newpage

### FPR by Trend Type

![Example trend functions](plots/seasonality_detection_trend_types_examples.pdf){#fig-trend-examples width="7in"}

![FPR across trend types and strengths](plots/seasonality_detection_trend_types_fpr.pdf){#fig-trend-types-fpr width="7in"}

\newpage

**How to interpret**:

- Each panel shows one trend type
- Lines show how FPR changes as trend strength increases
- **Slow sine is catastrophic for FFT**: 100% FPR because FFT detects the slow oscillation as "seasonality"

### Most Problematic Trend Types

| Trend Type | FFT FPR | Spectral FPR | Issue |
|------------|---------|--------------|-------|
| **slow_sine** | **100%** | 0% | FFT detects non-seasonal oscillation |
| quadratic | 10% | 5% | Minor |
| sigmoid | 5% | 5% | Minor |
| linear | 0% | 10% | Handled well |

# Key Findings

## Method Ranking

1. **Variance Strength** (F1=97.3%, FPR=2%): Best overall when period is known
2. **Spectral Strength** (F1=95.3%, FPR=10%): Most robust to different trend types
3. **FFT Confidence** (F1=94.8%, FPR=4%): Good but vulnerable to slow oscillations
4. **AIC Comparison** (F1=91.5%, FPR=18%): Interpretable but higher FPR
5. **ACF Confidence** (F1=85.4%, FPR=10%): Conservative, misses weak seasonality

## Critical Issues Found

1. **Period units matter**: The `period` parameter in `seasonal_strength()` must be in argvals units, not raw time units (e.g., 0.2 not 12)

2. **FFT is vulnerable to slow oscillations**: Any periodic signal (even non-seasonal) triggers detection

# Recommendations

## For Unknown Datasets

**Primary recommendation: Variance Strength**

```{r}
#| eval: false

# Calculate period in argvals units
# If argvals is in [0,1] and you expect 5 annual cycles:
period_in_argvals_units <- 1 / 5  # = 0.2

strength <- seasonal_strength(fd,
                              period = period_in_argvals_units,
                              method = "variance",
                              detrend = "linear")
is_seasonal <- strength > 0.2
```

**For robustness to unknown trends: Spectral Strength**

```{r}
#| eval: false

strength <- seasonal_strength(fd,
                              period = period_in_argvals_units,
                              method = "spectral",
                              detrend = "linear")
is_seasonal <- strength > 0.3
```

**Ensemble approach (most robust)**:

```{r}
#| eval: false

var_detected <- seasonal_strength(fd, period, method = "variance") > 0.2
spec_detected <- seasonal_strength(fd, period, method = "spectral") > 0.3
fft_detected <- estimate_period(fd, method = "fft")$confidence > 6.0

# Majority vote
is_seasonal <- (var_detected + spec_detected + fft_detected) >= 2
```

## Threshold Guidelines

| Method | Threshold | Calibration |
|--------|-----------|-------------|
| Variance Strength | 0.2 | 95th percentile of noise ~0.17 |
| Spectral Strength | 0.3 | 95th percentile of noise ~0.29 |
| FFT Confidence | 6.0 | 95th percentile of noise ~5.7 |
| ACF Confidence | 0.25 | 95th percentile of noise ~0.22 |
| AIC Difference | 0 | Fourier better → positive difference |

**Calibration methodology**: All thresholds were calibrated using pure noise data (seasonal strength = 0, no trend) by taking the 95th percentile of each method's score distribution. This ensures approximately 5% false positive rate on clean data. Note that FPR may increase when confounding trends are present (see Simulation 2 and 3 results).

# Conclusion

For detecting seasonality in functional time series:

1. **Variance Strength** is the most accurate method when the seasonal period is known
2. **Spectral Strength** is most robust to confounding trends and unknown oscillations
3. **FFT Confidence** works well but is vulnerable to slow non-seasonal oscillations
4. **AIC Comparison** provides an interpretable alternative but has higher false positive rates
5. **ACF Confidence** is conservative (low FPR) but misses weak seasonality

The key insight is that simple variance-based decomposition outperforms more complex spectral methods when properly configured with the correct period parameter.

# Appendix: Fourier vs P-spline Comparison

The AIC comparison method is based on the observation that Fourier bases naturally capture periodic patterns better than P-splines for seasonal data.

![Fourier vs P-spline AIC comparison](plots/seasonal_basis_comparison.pdf){#fig-basis-comparison width="7in"}

\newpage

**How to interpret**:

- Top left: Mean AIC for both methods across seasonal strengths
- Top right: AIC difference (Fourier - P-spline) showing crossover point
- Bottom left: Fourier win rate at each strength level
- Bottom right: Example curves at different seasonal strengths

# Appendix: File Listing

All simulation scripts and results are in `scripts/seasonal_simulation/`:

- `seasonality_detection_comparison.R` — Main comparison (Simulation 1)
- `seasonality_detection_with_trend.R` — Non-linear trend study (Simulation 2)
- `seasonality_detection_trend_types.R` — Multiple trend types (Simulation 3)
- `seasonal_basis_comparison.R` — Fourier vs P-spline AIC study

PDF outputs are in the `plots/` subfolder.
